---
title: 'flashtpx: Simulation Run 3'
author: "Wei Wang, Kushal K Dey"
date: "March 31, 2016"
output:
  tufte::tufte_handout:
    highlight: tango
    toc: true
---

## Overview

This is the third simulation run of *flashtpx* without the non-negative constraint. The main idea is to try and replicate the results for the `FitGoM()` in CountClust or `topics()` model in the maptpx package due to Matt Taddy.

Without the non-negative constraint, *flashtpx* is basically applying *flash* with the covariance matrix for the data estimated from the GoM or topic model fitting on the counts data. Here we apply *flashtpx* on the counts data generated from a chosen simulation design and then interpret the results and compare the results to the PMA model fitting.

## Simulation Design

We load the packages and the functions we need to perform the model.
```{r warning=FALSE, message=FALSE}
library(ashr)
library(irlba)
library(PMA)
source("../R/flash.R")
```

In the first two simulation runs, we considered $K=2$, here we consider the model with $K=3$.

Again as in first simulation run, we take samples in batches of size $200$. The first $200$ samples are from cluster 1, next $200$ from cluster $2$, next 200 from cluster $3$ and the final $200$ are proportionally allocated to the three clusters. More clear presentation of the omega matrix or membership matrix can be obtained from the Structure plot below.

```{r}
n.out <- 200
omega_sim <- rbind(cbind(rep(1,n.out), rep(0,n.out), rep(0, n.out)), cbind(rep(0,n.out), rep(1,n.out), rep(0, n.out)),  cbind(rep(0,n.out), rep(0,n.out), rep(1, n.out)), cbind(seq(0.3,0.4,length.out=n.out), seq(0.4,0.2,length.out=n.out), 1 - seq(0.3,0.4,length.out=n.out) - seq(0.4,0.2,length.out=n.out)));

K <- dim(omega_sim)[2]

par(mar=c(2,2,2,2))
barplot(t(omega_sim), 
        col = 2:(K+1),
        axisnames = F, space = 0, 
        border = NA, 
        main=paste("No. of clusters=", K),
        las=1, ylim = c(0,1), cex.axis=1.5,cex.main=1.4)
```

Assume there are $100$ genes we have observed. Lets see how the clusters look by defining the cluster frequency matrix below. 

```{r echo=TRUE, eval=TRUE}
freq <- rbind(c(0.1,0.2,rep(0.70/98,98)),c(rep(0.70/98,98), 0.1,0.2), c(rep(0.4/49,49), 0.1, 0.2, rep(0.3/49,49)));

str(freq)
```

We now generate the counts data given the above membership probability and the cluster distribution matrices.

```{r}
counts <- t( do.call(cbind,
                   lapply(1:dim(omega_sim)[1], 
                          function(x) 
                          rmultinom(1,1000,prob=omega_sim[x,]%*%freq))))
dim(counts)
```

We next fit a standard topic model with $K=2$. 

```{r}
topic.fit <- maptpx::topics(counts, K=3);
omega <- topic.fit$omega;
theta <- topic.fit$theta;

```

## Brief methods overview

Under the standard topic model, we have 

$$ c_{ng} \sim Poi (c_{n+} \sum_{k=1}^{K} \omega_{nk} \theta_{kg}) $$

Let us define 

$$ \lambda_{ng} = c_{n+} \sum_{k=1}^{K} \omega_{nk} \theta_{kg} $$

Under the normal model, if $\lambda_{ng}$ is large, we can assume

$$ c_{ng} \sim N( \lambda_{ng}, \lambda_{ng}) $$

which is equivalent to saying 

$$ c_{ng} = \lambda_{ng} + e_{ng} \hspace{1 cm} e_{ng} \sim N(0, \lambda_{ng}) $$

For applying *flash*, we first estimate the $\lambda$ in the variance using topic model estimate (we call this $\lambda$ to be $\lambda^{\star}$).

to guard against the odd possibility that for some $n$ and $g$, this estimate $\lambda^{\star}$ could be $0$, we replace such cases as of now with a small value $0.0001$.

In this example model we have considered $c_{n+}=1000$.

```{r}
lambda_star <- 1000* (omega %*% t(theta));
lambda_star[lambda_star==0] <- 0.0001;
dim(lambda_star)
```

## FLASH model fitting

*flash* then fits the model

$$ c_{ng} = \sum_{k=1}^{K} l_{nk} f_{kg} + e_{ng} \hspace{1 cm} e_{ng} \sim N(0, \lambda^{\star}_{ng}) $$

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, results='hide'}
g1 = suppressMessages(flash(counts,sigmae2_true = lambda_star))
f = g1$f
l = g1$l
res = counts - l%*%t(f)
# g_new = flash(res,nonnegative = TRUE,sigmae2_true = lambda)
g2 = suppressMessages(flash(res,sigmae2_true = lambda_star))
l = g2$l
f = g2$f
res = res - l%*%t(f)
g3 = suppressMessages(flash(res,sigmae2_true = lambda_star))
l = g3$l
f = g3$f
res = res - l%*%t(f)
g4 = suppressMessages(flash(res,sigmae2_true = lambda_star))
l = g4$l
f = g4$f
res = res - l%*%t(f)
g5 = suppressMessages(flash(res,sigmae2_true = lambda_star))
l = g5$l
f = g5$f
res = res - l%*%t(f)
g6 = suppressMessages(flash(res,sigmae2_true = lambda_star))
l = g6$l
f = g6$f
par(mfrow = c(2, 2))
par(cex = 0.6)
par(mar = c(3, 3, 0.8, 0.8), oma = c(1, 1, 1, 1))
plot(g1$l,main="1st loading")
plot(g1$f,main="1st factor")
plot(g2$l,main = "2nd loading")
plot(g2$f, main = "2nd factor")
plot(g3$l,main = "3rd loading")
plot(g3$f, main = "3rd factor")
plot(g4$l,main = "4th loading")
plot(g4$f, main = "4th factor")
plot(g5$l,main = "5th loading")
plot(g5$f, main = "5th factor")
plot(g6$l,main = "6th loading")
plot(g6$f, main = "6th factor")
```

## PMA model fitting

We compare the output of *flash* with that of the `PMD()` function of the package [PMA](https://cran.r-project.org/web/packages/PMA/PMA.pdf).
We use their default settings for shrinkage.

```{r}
out <- PMD(counts,K=6);

par(mfrow = c(2, 2))
par(cex = 0.6)
par(mar = c(3, 3, 0.8, 0.8), oma = c(1, 1, 1, 1))
plot(out$u[,1],main="1st loading")
plot(out$v[,1],main="1st factor")
plot(out$u[,2],main = "2nd loading")
plot(out$v[,2], main = "2nd factor")
plot(out$u[,3],main = "3rd loading")
plot(out$v[,3], main = "3rd factor")
plot(out$u[,4],main = "4th loading")
plot(out$v[,4], main = "4th factor")
plot(out$u[,5],main = "5th loading")
plot(out$v[,5], main = "5th factor")
plot(out$u[,6],main = "6th loading")
plot(out$v[,6], main = "6th factor")

```

```{r}
sessionInfo()
```

