---
title: "Importance of Variance weighting in flashtpx"
author: "Wei Wang, Kushal K Dey"
date: "April 1, 2016"
output:
  tufte::tufte_handout:
    highlight: tango
    toc: true
---

## Overview

In this script, we look at the significance of variance weighting in `flashtpx` models. Note that in the general normal topic model (NTM), we assume that each gene has the same variance across samples. If I consider this assumption under the refined normal model on counts, that would look like 

$$ FLASH 1: \hspace{1 cm} c_{ng} \sim N(\lambda_{ng}, \phi_{g}) $$

We may even make the assumption stricter and assume just a global variance $\phi$ instead of depending on the gene in $\phi_{g}$.

$$ FLASH 2: \hspace{1 cm} c_{ng} \sim N(\lambda_{ng}, \phi)  $$

 We want to compare these two models of *flash* with the one originally proposed in the simulation runs.

$$ FLASH 3: c_{ng} \sim N(\lambda_{ng}, \lambda_{ng}) $$

where the variance term $\lambda_{ng}$ is replaced by the estimate obtained from topic model $\lambda^{\star}_{ng}$ and then the *flash* model is performed for the mean $\lambda_{ng}$ with variance $\lambda^{\star}_{ng}$. 

Since the third model is more informative and closely resembles the Poisson model that is known to fit counts data well, we feel that it would do better than the first and second models. But is the performance comparable from the loadings and the factors we get? That is the question we try to answer using a simulation example and then running both these methods on that simulation data.

We use the same model as in Simulation Expt 1.

## Simulation Design

We load the packages and the functions we need to perform the model.
```{r warning=FALSE, message=FALSE}
library(ashr)
library(irlba)
library(PMA)
source("../R/flash.R")
```

Next we determine the `omega` matrix determined in batches of $200$, the first 200 batches of samples coming from one cluster completely, the next $200$ coming from a second cluster and the final $200$ being proportionally assigned to the two clusters. (Check barplot for more clarity).

```{r}
n.out <- 200
omega_sim <- rbind( cbind( rep(1, n.out), rep(0, n.out)), 
                  cbind( rep(0, n.out), rep(1, n.out)),
                  cbind( seq(0.6, 0.4, length.out = n.out), 
                         1- seq(0.6, 0.4,length.out=n.out)) )
dim(omega_sim)
K <- dim(omega_sim)[2]

par(mar=c(2,2,2,2))
barplot(t(omega_sim), 
        col = 2:(K+1),
        axisnames = F, space = 0, 
        border = NA, 
        main=paste("No. of clusters=", K),
        las=1, ylim = c(0,1), cex.axis=1.5,cex.main=1.4)

```

So we have two clusters. How do these clusters look? Assume we have $100$ genes.

```{r}
freq <- rbind( c(0.1, 0.2, rep(0.70/98, 98)),
             c(rep(0.70/98,98), 0.1, 0.2) )
str(freq)
```

So the first cluster has high expression at first 2 genes and low expression at the other 98 genes.

```{r}
counts <- t( do.call(cbind,
                   lapply(1:dim(omega_sim)[1], 
                          function(x) 
                          rmultinom(1,1000,prob=omega_sim[x,]%*%freq))))
dim(counts)
```

We first perform a topic model with $K=2$.

```{r}
topic.fit <- maptpx::topics(counts, K=2);
omega <- topic.fit$omega;
theta <- topic.fit$theta;
```

We obtain an estimate of the mean $\lambda_{ng}$ from the cluster membership probability (omega) matrix and the cluster distribution (theta) matrix. We then take the residuals of the counts from the mean matrix $\lambda$.

```{r}
lambda <- 1000* (omega %*% t(theta));
res_counts <- counts -lambda;
```


## FLASH model 1 fitting

*flash* then fits the model

$$ c_{ng} = \sum_{k=1}^{K} l_{nk} f_{kg} + e_{ng} \hspace{1 cm} e_{ng} \sim N(0, \phi_{g}) $$

We first determine the $\phi_{g}$ using the residual counts obtained after fitting topic model above and then computing the variance for each gene.

```{r}
var_genes <- apply(res_counts, 2, var);

rep.row<-function(x,n){
   matrix(rep(x,each=n),nrow=n)
}

phi <- rep.row(var_genes, dim(res_counts)[1])
phi[phi==0] <- 0.0001
dim(phi)
```


```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, results='hide'}
g1 = suppressMessages(flash(counts,sigmae2_true = phi))
f = g1$f
l = g1$l
res = counts - l%*%t(f)
# g_new = flash(res,nonnegative = TRUE,sigmae2_true = lambda)
g2 = suppressMessages(flash(res,sigmae2_true = phi))
l = g2$l
f = g2$f
res = res - l%*%t(f)
g3 = suppressMessages(flash(res,sigmae2_true = phi))
l = g3$l
f = g3$f
res = res - l%*%t(f)
g4 = suppressMessages(flash(res,sigmae2_true = phi))
l = g4$l
f = g4$f
res = res - l%*%t(f)
g5 = suppressMessages(flash(res,sigmae2_true = phi))
l = g5$l
f = g5$f
res = res - l%*%t(f)
g6 = suppressMessages(flash(res,sigmae2_true = phi))
l = g6$l
f = g6$f
par(mfrow = c(2, 2))
par(cex = 0.6)
par(mar = c(3, 3, 0.8, 0.8), oma = c(1, 1, 1, 1))
plot(g1$l,main="1st loading")
plot(g1$f,main="1st factor")
plot(g2$l,main = "2nd loading")
plot(g2$f, main = "2nd factor")
plot(g3$l,main = "3rd loading")
plot(g3$f, main = "3rd factor")
plot(g4$l,main = "4th loading")
plot(g4$f, main = "4th factor")
plot(g5$l,main = "5th loading")
plot(g5$f, main = "5th factor")
plot(g6$l,main = "6th loading")
plot(g6$f, main = "6th factor")
```

## FLASH model 2 fitting

*flash* then fits the model

$$ c_{ng} = \sum_{k=1}^{K} l_{nk} f_{kg} + e_{ng} \hspace{1 cm} e_{ng} \sim N(0, \phi) $$

We first determine the $\phi$ using the residual counts obtained after fitting topic model above and then computing the variance of all the residuals.

```{r}
var_total <- var(as.vector(res_counts));

phi <- matrix(var_total, nrow=dim(res_counts)[1], ncol=dim(res_counts)[2]);
phi[phi==0] <- 0.0001
dim(phi)
```


```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, results='hide'}
g1 = suppressMessages(flash(counts,sigmae2_true = phi))
f = g1$f
l = g1$l
res = counts - l%*%t(f)
# g_new = flash(res,nonnegative = TRUE,sigmae2_true = lambda)
g2 = suppressMessages(flash(res,sigmae2_true = phi))
l = g2$l
f = g2$f
res = res - l%*%t(f)
g3 = suppressMessages(flash(res,sigmae2_true = phi))
l = g3$l
f = g3$f
res = res - l%*%t(f)
g4 = suppressMessages(flash(res,sigmae2_true = phi))
l = g4$l
f = g4$f
res = res - l%*%t(f)
g5 = suppressMessages(flash(res,sigmae2_true = phi))
l = g5$l
f = g5$f
res = res - l%*%t(f)
g6 = suppressMessages(flash(res,sigmae2_true = phi))
l = g6$l
f = g6$f
par(mfrow = c(2, 2))
par(cex = 0.6)
par(mar = c(3, 3, 0.8, 0.8), oma = c(1, 1, 1, 1))
plot(g1$l,main="1st loading")
plot(g1$f,main="1st factor")
plot(g2$l,main = "2nd loading")
plot(g2$f, main = "2nd factor")
plot(g3$l,main = "3rd loading")
plot(g3$f, main = "3rd factor")
plot(g4$l,main = "4th loading")
plot(g4$f, main = "4th factor")
plot(g5$l,main = "5th loading")
plot(g5$f, main = "5th factor")
plot(g6$l,main = "6th loading")
plot(g6$f, main = "6th factor")
```

## FLASH model 3 fitting

*flash* then fits the model

$$ c_{ng} = \sum_{k=1}^{K} l_{nk} f_{kg} + e_{ng} \hspace{1 cm} e_{ng} \sim N(0, \lambda^{\star}_{ng}) $$

where we ontain $\lambda^{\star}_{ng}$ as follows

```{r}
lambda_star <- 1000* (omega %*% t(theta));
lambda_star[lambda_star==0] <- 0.0001;
dim(lambda_star)
```

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, results='hide'}
g1 = suppressMessages(flash(counts,sigmae2_true = lambda_star))
f = g1$f
l = g1$l
res = counts - l%*%t(f)
# g_new = flash(res,nonnegative = TRUE,sigmae2_true = lambda)
g2 = suppressMessages(flash(res,sigmae2_true = lambda_star))
l = g2$l
f = g2$f
res = res - l%*%t(f)
g3 = suppressMessages(flash(res,sigmae2_true = lambda_star))
l = g3$l
f = g3$f
res = res - l%*%t(f)
g4 = suppressMessages(flash(res,sigmae2_true = lambda_star))
l = g4$l
f = g4$f
res = res - l%*%t(f)
g5 = suppressMessages(flash(res,sigmae2_true = lambda_star))
l = g5$l
f = g5$f
res = res - l%*%t(f)
g6 = suppressMessages(flash(res,sigmae2_true = lambda_star))
l = g6$l
f = g6$f
par(mfrow = c(2, 2))
par(cex = 0.6)
par(mar = c(3, 3, 0.8, 0.8), oma = c(1, 1, 1, 1))
plot(g1$l,main="1st loading")
plot(g1$f,main="1st factor")
plot(g2$l,main = "2nd loading")
plot(g2$f, main = "2nd factor")
plot(g3$l,main = "3rd loading")
plot(g3$f, main = "3rd factor")
plot(g4$l,main = "4th loading")
plot(g4$f, main = "4th factor")
plot(g5$l,main = "5th loading")
plot(g5$f, main = "5th factor")
plot(g6$l,main = "6th loading")
plot(g6$f, main = "6th factor")
```

```{r}
sessionInfo()
```

