---
title: 'Real Data applicaion of flashtpx: Bird Abundance Data due to Price et al'
author: "Wei Wang, Kushal K Dey"
date: "April 1, 2016"
output:
  tufte::tufte_handout:
    highlight: tango
    toc: true
---

## Overview

We now apply `flashtpx` on a real data. This data is collected by Trevor Price et al and does not yet have open access. But the good thing about this data is that $K=2$ and $K=3$ clusters, the standard GoM model gave pretty interesting results where it seemed to pick up the elevation gradient for $K=2$ and the geographic gradient (East and West) for $K=3$. The other good thing is that the dataset is pretty small and since there is an issue about the scaling of `flashtpx` for large datasets, this seems like the ideal first real data to apply this method on.

We also compare the results from standard goM model fit, the `flashtpx` model fit and the PMA model fit.

## Data Preparation

```{r warning=FALSE, message=FALSE}
rm(list=ls())
library(ashr)
library(irlba)
library(PMA)
library(maptpx)
source("../R/flash.R")
```


```{r}
data=read.csv("../external_data/Himalayan_grid_matrix.csv",header=TRUE);

counts=as.matrix(data[,-1]);

rownames(counts)=data[,1];

new_data1 <- data.frame(read.csv('../external_data/MohanGrids.csv'));
new_data2 <- data.frame(read.csv('../external_data/MohanGrids2.csv'));

bird_species <- union(as.vector(colnames(counts)), union(as.vector(new_data1[,1]), as.vector(new_data2[,1])));

new_data <- matrix(0,dim(counts)[1]+3,length(bird_species));
new_data[1,match(new_data1[,1],bird_species)]=new_data1[,2];
new_data[2,match(new_data1[,1],bird_species)]=new_data1[,3];
new_data[3,match(new_data2[,1],bird_species)]=new_data2[,2];
new_data[4:(dim(counts)[1]+3),match(colnames(counts),bird_species)]=counts;

new_counts <- as.matrix(new_data);

rownames(new_counts) <- c(c("U1","U2","MA1"),rownames(counts));
colnames(new_counts) <- bird_species;
new_counts <- new_counts[-(1:3),];
```

Next we load the metadata on the samples.

```{r}
metadata=read.csv("../external_data/Himalayan_grid_metadata.csv",header=TRUE);
elevation_metadata=metadata$Elevation[match(rownames(new_counts),metadata[,1])];
east_west_dir = metadata$WorE[match(rownames(new_counts),metadata[,1])];

```

## Standard GoM model fitting

We now fit the topic model for $K=2$

```{r}
Topic_clus <- maptpx::topics(new_counts, K=2)
```

```{r echo=TRUE, eval=TRUE}
K <- 2
par(mfrow=c(1,1))
par(mar=c(8,2,2,2))
barplot(t(Topic_clus$omega[order(elevation_metadata),]),col=2:(K+1),
        axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),
        cex.axis=0.2,cex.main=1.4, xlab="elevation")
title(main=paste("Structure Plot topic proportions,k=",K))
combo_patch_dir = paste0(round(elevation_metadata));
combo_patch_dir_ordered = combo_patch_dir[order(elevation_metadata)];

match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],35);
match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);

axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
abline(v=match_labs[2:length(match_labs)]);

```

We perform the topic model for $K=3$.

```{r}
Topic_clus <- maptpx::topics(new_counts, K=3)
```

```{r echo=TRUE, eval=TRUE}

docweights <- Topic_clus$omega
par(mfrow=c(1,1))
par(mar=c(8,2,2,2))
east_west_elevation = paste0(metadata$WorE, "_", metadata$Elevation);

index1 <- which(metadata$WorE=="E");
index2 <- which(metadata$WorE=="W");
elevation1 <- metadata$Elevation[index1]; elevation2 <- metadata$Elevation[index2];
index_WE <- c(index1[order(elevation1)], index2[order(elevation2)]);

  K <- 3
    barplot(t(docweights[index_WE,]),col=2:(K+1),axisnames=F,
            space=0,border=NA,main=paste("No. of clusters=",3),las=1,ylim=c(0,1),cex.axis=0.2,cex.main=1.4)

    combo_patch_dir = paste0(east_west_elevation);
    combo_patch_dir_ordered = combo_patch_dir[index_WE];

  match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
  match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],35);
  match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
  labs=match_labs_prefix - 0.5*(match_labs_suffix - match_labs_prefix);

  axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
  abline(v=match_labs[1:length(match_labs)]);
```

## FLASH model fitting

We first derive the varaince weights.

```{r}
lambda_star <- 1000* (Topic_clus$omega %*% t(Topic_clus$theta));
lambda_star[lambda_star==0] <- 0.0001;
dim(lambda_star)
```

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, results='hide'}
g1 = suppressMessages(flash(new_counts,sigmae2_true = lambda_star))
f = g1$f
l = g1$l
res = new_counts - l%*%t(f)
# g_new = flash(res,nonnegative = TRUE,sigmae2_true = lambda)
g2 = suppressMessages(flash(res,sigmae2_true = lambda_star))
l = g2$l
f = g2$f
res = res - l%*%t(f)
g3 = suppressMessages(flash(res,sigmae2_true = lambda_star))
l = g3$l
f = g3$f
res = res - l%*%t(f)
g4 = suppressMessages(flash(res,sigmae2_true = lambda_star))
l = g4$l
f = g4$f
res = res - l%*%t(f)
g5 = suppressMessages(flash(res,sigmae2_true = lambda_star))
l = g5$l
f = g5$f
res = res - l%*%t(f)
g6 = suppressMessages(flash(res,sigmae2_true = lambda_star))
l = g6$l
f = g6$f
```

We plot the factor loadings in the order of elevation as in the structure plot for the GoM model with $K=2$.

```{r}
par(mfrow = c(2, 2))
par(cex = 0.6)
par(mar = c(3, 3, 0.8, 0.8), oma = c(1, 1, 1, 1))
plot(g1$l[order(elevation_metadata)],main="1st loading")
plot(g1$f,main="1st factor")
plot(g2$l[order(elevation_metadata)],main = "2nd loading")
plot(g2$f, main = "2nd factor")
plot(g3$l[order(elevation_metadata)],main = "3rd loading")
plot(g3$f, main = "3rd factor")
plot(g4$l[order(elevation_metadata)],main = "4th loading")
plot(g4$f, main = "4th factor")
plot(g5$l[order(elevation_metadata)],main = "5th loading")
plot(g5$f, main = "5th factor")
plot(g6$l[order(elevation_metadata)],main = "6th loading")
plot(g6$f, main = "6th factor")
```

We plot the factor loadings in the order of both geography and elevation as in the structure plot for the GoM model with $K=3$.

```{r}
par(mfrow = c(2, 2))
par(cex = 0.6)
par(mar = c(3, 3, 0.8, 0.8), oma = c(1, 1, 1, 1))
plot(g1$l[index_WE],main="1st loading")
plot(g1$f,main="1st factor")
plot(g2$l[index_WE],main = "2nd loading")
plot(g2$f, main = "2nd factor")
plot(g3$l[index_WE],main = "3rd loading")
plot(g3$f, main = "3rd factor")
plot(g4$l[index_WE],main = "4th loading")
plot(g4$f, main = "4th factor")
plot(g5$l[index_WE],main = "5th loading")
plot(g5$f, main = "5th factor")
plot(g6$l[index_WE],main = "6th loading")
plot(g6$f, main = "6th factor")
```

## PMA model fitting

```{r}
out <- PMD(new_counts,K=6);
```

The loadings and factors ordered in the same order as the Structure plot for $K=2$.

```{r}
par(mfrow = c(2, 2))
par(cex = 0.6)
par(mar = c(3, 3, 0.8, 0.8), oma = c(1, 1, 1, 1))
plot(out$u[order(elevation_metadata),1],main="1st loading")
plot(out$v[,1],main="1st factor")
plot(out$u[order(elevation_metadata),2],main = "2nd loading")
plot(out$v[,2], main = "2nd factor")
plot(out$u[order(elevation_metadata),3],main = "3rd loading")
plot(out$v[,3], main = "3rd factor")
plot(out$u[order(elevation_metadata),4],main = "4th loading")
plot(out$v[,4], main = "4th factor")
plot(out$u[order(elevation_metadata),5],main = "5th loading")
plot(out$v[,5], main = "5th factor")
plot(out$u[order(elevation_metadata),6],main = "6th loading")
plot(out$v[,6], main = "6th factor")

```

The loadings and factors ordered in the same order as the Structure plot for $K=3$.

```{r}
par(mfrow = c(2, 2))
par(cex = 0.6)
par(mar = c(3, 3, 0.8, 0.8), oma = c(1, 1, 1, 1))
plot(out$u[index_WE,1],main="1st loading")
plot(out$v[,1],main="1st factor")
plot(out$u[index_WE,2],main = "2nd loading")
plot(out$v[,2], main = "2nd factor")
plot(out$u[index_WE,3],main = "3rd loading")
plot(out$v[,3], main = "3rd factor")
plot(out$u[index_WE,4],main = "4th loading")
plot(out$v[,4], main = "4th factor")
plot(out$u[index_WE,5],main = "5th loading")
plot(out$v[,5], main = "5th factor")
plot(out$u[index_WE,6],main = "6th loading")
plot(out$v[,6], main = "6th factor")

```
