---
title: "Factor analysis to replicate Topic model results"
author: "Kushal K Dey"
date: "April 9, 2016"
output:
  tufte::tufte_handout:
    highlight: tango
    toc: true
---

## Overview

The package PMA due to Witten and Tibshirani is a very popular matrix decomposition package. In this script, we check if we use a variance scaling transform using Cholesky decomposition that sort of tries to pool the Poisson model features.

## Simulation experiment 1

```{r}
n.out <- 800
omega_sim <-  cbind(seq(0.6,0.4,length.out=n.out), 
                    1- seq(0.6,0.4,length.out=n.out) )
colSums(omega_sim)

K <- dim(omega_sim)[2]
barplot(t(omega_sim),col=2:(K+1),
        axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,
        ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

freq <- rbind( c(0.1, 0.2, rep(0.70/98,98)),
               c(rep(0.70/98,98), 0.1, 0.2))

counts <- t(do.call(cbind,lapply(1:dim(omega_sim)[1], function(x) rmultinom(1,1000,prob=omega_sim[x,]%*%freq))));

lambda <- 1000*(omega_sim%*%freq);
lambda[lambda==0] <- 0.0001;
dim(lambda)

scaled_counts <- counts/sqrt(lambda);

require(PMA)
out <- PMD(scaled_counts, K=4, upos=TRUE, vpos=TRUE, center=TRUE,
            sumabs=1, niter=20000, sumabsu=sqrt(600))

plot(out$u[,1],main="1st loading")
plot(out$v[,1],main="1st factor")
plot(out$u[,2],main = "2nd loading")
plot(out$v[,2], main = "2nd factor")
plot(out$u[,3],main = "3rd loading")
plot(out$v[,3], main = "3rd factor")

omega1 <- maptpx::normalize(cbind(out$u[,2], out$u[,3]), byrow=TRUE);

barplot(t(omega1),col=2:(K+1),
        axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,
        ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

omega2 <- maptpx::normalize(cbind(out$u[,3], out$u[,4]), byrow=TRUE);

barplot(t(omega2),col=2:(K+1),
        axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,
        ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

omega3 <- maptpx::normalize(cbind(out$u[,2], out$u[,4]), byrow=TRUE);

barplot(t(omega3),col=2:(K+1),
        axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,
        ylim=c(0,1),cex.axis=1.5,cex.main=1.4)


omega4 <- maptpx::normalize(cbind(out$u[,2], out$u[,3], out$u[,4]), byrow=TRUE);

barplot(t(omega4),col=2:(K+2),
        axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,
        ylim=c(0,1),cex.axis=1.5,cex.main=1.4)


tpx.fit <- maptpx::topics(counts, K=3)
barplot(t(tpx.fit$omega),col=2:(K+2),
        axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,
        ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

tpx.fit <- maptpx::topics(counts, K=2)
barplot(t(tpx.fit$omega),col=2:(K+2),
        axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,
        ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
```


## Simulation experiment 2

```{r}
n.out <- 200
omega_sim <- rbind( cbind( rep(1, n.out), rep(0, n.out)), 
                  cbind( rep(0, n.out), rep(1, n.out)),
                  cbind( seq(0.6, 0.4, length.out = n.out), 
                         1- seq(0.6, 0.4,length.out=n.out)) )
dim(omega_sim)
K <- dim(omega_sim)[2]

par(mar=c(2,2,2,2))
barplot(t(omega_sim), 
        col = 2:(K+1),
        axisnames = F, space = 0, 
        border = NA, 
        main=paste("No. of clusters=", K),
        las=1, ylim = c(0,1), cex.axis=1.5,cex.main=1.4)

freq <- rbind( c(0.1, 0.2, rep(0.70/98, 98)),
             c(rep(0.70/98,98), 0.1, 0.2) )
str(freq)

counts <- t( do.call(cbind,
                   lapply(1:dim(omega_sim)[1], 
                          function(x) 
                          rmultinom(1,1000,prob=omega_sim[x,]%*%freq))))
dim(counts)

lambda <- 1000*(omega_sim%*%freq);
lambda[lambda==0] <- 0.0001;
dim(lambda)

scaled_counts <- counts/sqrt(lambda);

require(PMA)
out <- PMD(scaled_counts, K=4, upos=TRUE, vpos=TRUE, center=TRUE,
            sumabs=1, niter=20000, sumabsu=sqrt(600))

plot(out$u[,1],main="1st loading")
plot(out$v[,1],main="1st factor")
plot(out$u[,2],main = "2nd loading")
plot(out$v[,2], main = "2nd factor")
plot(out$u[,3],main = "3rd loading")
plot(out$v[,3], main = "3rd factor")

omega1 <- maptpx::normalize(cbind(out$u[,2], out$u[,3]), byrow=TRUE);

barplot(t(omega1),col=2:(K+1),
        axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,
        ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

omega2 <- maptpx::normalize(cbind(out$u[,3], out$u[,4]), byrow=TRUE);

barplot(t(omega2),col=2:(K+1),
        axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,
        ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

omega3 <- maptpx::normalize(cbind(out$u[,2], out$u[,4]), byrow=TRUE);

barplot(t(omega3),col=2:(K+1),
        axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,
        ylim=c(0,1),cex.axis=1.5,cex.main=1.4)


omega4 <- maptpx::normalize(cbind(out$u[,2], out$u[,3], out$u[,4]), byrow=TRUE);

barplot(t(omega4),col=2:(K+2),
        axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,
        ylim=c(0,1),cex.axis=1.5,cex.main=1.4)


tpx.fit <- maptpx::topics(counts, K=3)
barplot(t(tpx.fit$omega),col=2:(K+2),
        axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,
        ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

tpx.fit <- maptpx::topics(counts, K=2)
barplot(t(tpx.fit$omega),col=2:(K+1),
        axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,
        ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
```

## Simulation experiment 3 

This experiment is similar to Simulation model 1 but with lot more samples and genes.

```{r}
n.out <- 1000
omega_sim <- rbind( cbind( rep(1, n.out), rep(0, n.out)), 
                    cbind( rep(0, n.out), rep(1, n.out)),
                    cbind( seq(0.6, 0.4, length.out = n.out), 
                           1- seq(0.6, 0.4,length.out=n.out)) )
dim(omega_sim)

K <- dim(omega_sim)[2]

par(mar=c(2,2,2,2))
barplot(t(omega_sim), 
        col = 2:(K+1),
        axisnames = F, space = 0, 
        border = NA, 
        main=paste("No. of clusters=", K),
        las=1, ylim = c(0,1), cex.axis=1.5,cex.main=1.4)

freq <- rbind( c(0.1, 0.2, rep(0.70/998, 998)),
               c(rep(0.70/998,998), 0.1, 0.2) )
str(freq)

counts <- t( do.call(cbind,
                   lapply(1:dim(omega_sim)[1], 
                          function(x) 
                          rmultinom(1,1000,prob=omega_sim[x,]%*%freq))))
dim(counts)

lambda <- 1000*(omega_sim%*%freq);
lambda[lambda==0] <- 0.0001;
dim(lambda)


scaled_counts <- counts/sqrt(lambda);

require(PMA)
out <- PMD(scaled_counts, K=4, upos=TRUE, vpos=TRUE, center=TRUE,
            sumabs=1, niter=20000, sumabsu=sqrt(600))

plot(out$u[,1],main="1st loading")
plot(out$v[,1],main="1st factor")
plot(out$u[,2],main = "2nd loading")
plot(out$v[,2], main = "2nd factor")
plot(out$u[,3],main = "3rd loading")
plot(out$v[,3], main = "3rd factor")

omega1 <- maptpx::normalize(cbind(out$u[,2], out$u[,3]), byrow=TRUE);

barplot(t(omega1),col=2:(K+1),
        axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,
        ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

omega2 <- maptpx::normalize(cbind(out$u[,3], out$u[,4]), byrow=TRUE);

barplot(t(omega2),col=2:(K+1),
        axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,
        ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

omega3 <- maptpx::normalize(cbind(out$u[,2], out$u[,4]), byrow=TRUE);

barplot(t(omega3),col=2:(K+1),
        axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,
        ylim=c(0,1),cex.axis=1.5,cex.main=1.4)


omega4 <- maptpx::normalize(cbind(out$u[,2], out$u[,3], out$u[,4]), byrow=TRUE);

barplot(t(omega4),col=2:(K+2),
        axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,
        ylim=c(0,1),cex.axis=1.5,cex.main=1.4)


#tpx.fit <- maptpx::topics(counts, K=3)
#save(tpx.fit, file="../rdas/pma_tpx_compare_1.rda")

tpx.fit <- get(load(file="../rdas/pma_tpx_compare_1.rda"))
barplot(t(tpx.fit$omega),col=2:(K+2),
        axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,
        ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

tpx.fit <- maptpx::topics(counts, K=2)
barplot(t(tpx.fit$omega),col=2:(K+1),
        axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,
        ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
```

## Simulation Experiment 4

```{r}
n.out <- 200
omega_sim <- rbind(cbind(rep(1,n.out), rep(0,n.out), rep(0, n.out)), cbind(rep(0,n.out), rep(1,n.out), rep(0, n.out)),  cbind(rep(0,n.out), rep(0,n.out), rep(1, n.out)), cbind(seq(0.3,0.4,length.out=n.out), seq(0.4,0.2,length.out=n.out), 1 - seq(0.3,0.4,length.out=n.out) - seq(0.4,0.2,length.out=n.out)));

K <- dim(omega_sim)[2]

par(mar=c(2,2,2,2))
barplot(t(omega_sim), 
        col = 2:(K+1),
        axisnames = F, space = 0, 
        border = NA, 
        main=paste("No. of clusters=", K),
        las=1, ylim = c(0,1), cex.axis=1.5,cex.main=1.4)

freq <- rbind(c(0.1,0.2,rep(0.70/98,98)),c(rep(0.70/98,98), 0.1,0.2), c(rep(0.4/49,49), 0.1, 0.2, rep(0.3/49,49)));

str(freq)

counts <- t( do.call(cbind,
                   lapply(1:dim(omega_sim)[1], 
                          function(x) 
                          rmultinom(1,1000,prob=omega_sim[x,]%*%freq))))
dim(counts)

lambda <- 1000*(omega_sim%*%freq);
lambda[lambda==0] <- 0.0001;
dim(lambda)


scaled_counts <- counts/sqrt(lambda);

require(PMA)
out <- PMD(scaled_counts, K=5, upos=TRUE, vpos=TRUE, center=TRUE,
            sumabs=1, niter=20000, sumabsu=sqrt(600))

plot(out$u[,1],main="1st loading")
plot(out$v[,1],main="1st factor")
plot(out$u[,2],main = "2nd loading")
plot(out$v[,2], main = "2nd factor")
plot(out$u[,3],main = "3rd loading")
plot(out$v[,3], main = "3rd factor")

K <- 3
omega1 <- maptpx::normalize(cbind(out$u[,2], out$u[,3], out$u[,4]), byrow=TRUE);

barplot(t(omega1),col=2:(K+1),
        axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,
        ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

omega2 <- maptpx::normalize(cbind(out$u[,3], out$u[,4], out$u[,5]), byrow=TRUE);

barplot(t(omega2),col=2:(K+1),
        axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,
        ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

omega3 <- maptpx::normalize(cbind(out$u[,2], out$u[,4], out$u[,5]), byrow=TRUE);

barplot(t(omega3),col=2:(K+1),
        axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,
        ylim=c(0,1),cex.axis=1.5,cex.main=1.4)


tpx.fit <- maptpx::topics(counts, K=3)
barplot(t(tpx.fit$omega),col=2:(K+2),
        axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,
        ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

```

